Distributional Trees and Forests

### 2017-10-05 pkg/disttree/inst/draft/draft.Rnw
adapted to the structure of the results from rain_cross and rain_pred
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1622 255f4249-2263-4417-b1ea-8498779e27a7

## function to estimate standard deviation of randomForest for a new observation
# (in randomForest the argument 'keep.inbag' must be set to TRUE)
rf_getsd <- function(rf, newdata, rfdata){

@@

\author{Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}

@@

  Achim Zeileis \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Achim.Zeileis@R-project.org} \\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}

@@

<<echo=FALSE, results=hide>>=
source("onecov.R")
if(file.exists("onecov1.rda")){
  load("onecov1.rda")
} else {
  onecov1 <- sim_onecov(kappa = 1, nobs = 700,
                         seedconst = 7, ntree = 100,
                         formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
                         tree_minsplit = 40, tree_mincrit = 0.95,
                         forest_minsplit = 20, forest_mincrit = 0, 
                         type.tree = "ctree",
                         mubase = 0,
                         censNO = TRUE,
                         fix.mu = FALSE,
                         fix.sigma = FALSE,
                         mu.sigma.interaction = FALSE,
                         gamboost_cvr = FALSE,
                         eval_disttree = TRUE,
                         eval_distforest = TRUE,
                         eval_gamlss = TRUE,
                         eval_bamlss = TRUE,
                         eval_gamboostLSS = TRUE,
                         eval_randomForest = TRUE,
                         eval_cforest = TRUE)
  save(onecov1, file = "onecov1.rda")
}

if(file.exists("onecov10.rda")){
  load("onecov10.rda")
} else {
  onecov10 <- sim_onecov(kappa = 10, nobs = 700,
                         seedconst = 7, ntree = 100,
                         formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
                         tree_minsplit = 40, tree_mincrit = 0.95,
                         forest_minsplit = 20, forest_mincrit = 0, 
                         type.tree = "ctree",
                         mubase = 0,
                         censNO = TRUE,
                         fix.mu = FALSE,
                         fix.sigma = FALSE,
                         mu.sigma.interaction = FALSE,
                         gamboost_cvr = FALSE,
                         eval_disttree = TRUE,
                         eval_distforest = TRUE,
                         eval_gamlss = TRUE,
                         eval_bamlss = TRUE,
                         eval_gamboostLSS = TRUE,
                         eval_randomForest = TRUE,
                         eval_cforest = TRUE)
  save(onecov10, file = "onecov10.rda")
}
@

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<sim_onecov1_mu, fig=TRUE, echo=FALSE>>=
plot_onecov(learndata = onecov1$learndata, parfun = onecov1$fun, 
            dt = onecov1$dt, df = onecov1$df, g = onecov1$g, b = onecov1$b, 
            gb = onecov1$gb, rf = onecov1$rf, cf = onecov1$cf, 
            compare_mu = TRUE, fix_x2 = 0,
            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<sim_onecov1_sigma, fig=TRUE, echo=FALSE>>=
plot_onecov(learndata = onecov1$learndata, parfun = onecov1$fun, 
            dt = onecov1$dt, df = onecov1$df, g = onecov1$g, b = onecov1$b, 
            gb = onecov1$gb, rf = onecov1$rf, cf = onecov1$cf, 
            compare_sigma = TRUE, fix_x1 = 0.4,
            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
@
\end{subfigure}
\caption{\label{plot_sim_onecov1} ... }
\end{figure}

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<sim_onecov10_mu, fig=TRUE, echo=FALSE>>=
plot_onecov(learndata = onecov10$learndata, parfun = onecov10$fun, 
            dt = onecov10$dt, df = onecov10$df, g = onecov10$g, b = onecov10$b, 
            gb = onecov10$gb, rf = onecov10$rf, cf = onecov10$cf, 
            compare_mu = TRUE, fix_x2 = 0,
            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<sim_onecov10_sigma, fig=TRUE, echo=FALSE>>=
plot_onecov(learndata = onecov10$learndata, parfun = onecov10$fun, 
            dt = onecov10$dt, df = onecov10$df, g = onecov10$g, b = onecov10$b, 
            gb = onecov10$gb, rf = onecov10$rf, cf = onecov10$cf, 
            compare_sigma = TRUE, fix_x1 = 0.4,
            add_dt = TRUE, add_df = TRUE, add_g = TRUE, add_b = TRUE, add_gb = TRUE, add_rf = TRUE, add_cf = TRUE)
@
\end{subfigure}
\caption{\label{plot_sim_onecov10} ... }
\end{figure}

@@

<<echo=FALSE, results=hide>>=
if(file.exists("simres.rda")){
  load("simres.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/inst/draft/.R")
  source("gensim.R")
  simres <- gensim(seedconst = 7, nrep = 40, ntree = 100,
                   nsteps = 9, stepsize = 1, kappa.start = 1,
                   formula = y~x1+x2+x3+x4+x5+x6+x7+x8+x9+x10, 
                   nobs = 400, testnobs = 200L,
                   tree_minsplit = 14, tree_mincrit = 0.95,
                   forest_minsplit = 7, forest_mincrit = 0, 
                   type.tree = "ctree",
                   censNO = TRUE,
                   fix.mu = FALSE,
                   fix.sigma = FALSE,
                   mu.sigma.interaction = FALSE,
                   gamboost_cvr = FALSE,
                   eval_disttree = TRUE,
                   eval_distforest = TRUE,
                   eval_gamlss = TRUE,
                   eval_bamlss = TRUE,
                   eval_gamboostLSS = TRUE,
                   eval_randomForest = TRUE,
                   eval_cforest = TRUE,
                   mubase = 0)
  save(simres, file = "simres.rda")
}
@

@@

ncuts <- 100
colR <- colorRampPalette(c("blue","green"))                                       
cols <- colR(ncuts+1)
griddata <- expand.grid(list(x1 = seq(-0.4, 1, 0.01), x2 = seq(-10, 10, 0.2)))
griddata$mu <- levelfun(griddata)
levelplot(mu ~ x1 * x2, data = griddata, xlab = "x1", ylab = "x2", 
          region = TRUE, cuts = ncuts, col.regions = cols)

@@

\\
%Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups. This is the point where the tree structure provides a better solution. However, a single tree can not keep up with the performance of \code{gamlss} when it comes to modeling the smooth functions within the subgroups. Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
%However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
%\\
%Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation. For $\kappa=1$ the performance of all three methods is similar. However, as $\kappa$ increases the RMSE of gamlss increases while the RMSE of trees and forests stays at almost the same level or even decreases sligthly as soon as $\kappa$ is greater or equal 2. The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
%This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
%For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become. Especially regarding the scale parameter the trees are clearly advantageous over the forests.

@@

\section{Rain}
In this section precipation data of .. years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
The following methods are applied on this data set:
\begin{itemize}
\item distributional tree
\item distributional forest
\item gamlss
\item gamboostLSS
\item bamlss
\item EMOS
\end{itemize}

@@

<<echo=FALSE, results=hide>>=
#### cross validation rain
if(file.exists("rainres.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rainres.rda")
  load("rainres.rda")
} else {
  source("~/svn/partykit/pkg/disttree/inst/draft/rain.R")
  #source("gensim.R")
  rainres <- rain(seedconst = 7423)
  save(rainres, file = "rainres.rda")
}

rain_rmse <- matrix(0, ncol = ncol(rainres[[1]]$rmse), length(rainres))
rain_ll <- matrix(0, ncol = ncol(rainres[[1]]$ll), length(rainres))
rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres))

for(i in 1:length(rainres)){
  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse)
  rain_ll[i,] <- colMeans(rainres[[i]]$ll)
  rain_crps[i,] <- colMeans(rainres[[i]]$crps)
}

colnames(rain_rmse) <- colnames(rain_ll) <- colnames(rain_crps) <-
  colnames(rainres[[1]]$rmse)
@

\begin{figure}[t!]
\begin{subfigure}{0.3\textwidth}
<<rain_rmse_box, fig=TRUE, echo=FALSE>>=
boxplot(rain_rmse)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.3\textwidth}
<<rain_ll_box, fig=TRUE, echo=FALSE>>=
boxplot(rain_ll)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.3\textwidth}
<<rain_crps_box, fig=TRUE, echo=FALSE>>=
boxplot(rain_crps)
@
\end{subfigure}
\caption{\label{boxplots_rain}Results 10x10 cross validation on the rain data set (RMSE, log-likelihood and CRPS)\\
left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
right: continuous lines represent the location parameter, dotted lines the scale parameter}
\end{figure}

ncuts <- 100
colR <- colorRampPalette(c("blue","green"))                                       
cols <- colR(ncuts+1)
griddata <- expand.grid(list(x1 = seq(-0.4, 1, 0.01), x2 = seq(-10, 10, 0.2)))
griddata$mu <- levelfun(griddata)
levelplot(mu ~ x1 * x2, data = griddata, xlab = "x1", ylab = "x2", 
          region = TRUE, cuts = ncuts, col.regions = cols)

>>>>

## function to estimate standard deviation of randomForest for a new observation
# (in randomForest the argument 'keep.inbag' must be set to TRUE)
rf_getsd <- function(rf, newdata = NULL, rfdata){

  if(is.null(newdata)) newdata <- rfdata

@@

\author{Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \And Achim Zeileis\\Universit\"at Innsbruck
   \And Reto Stauffer\\Universit\"at Innsbruck}
\Plainauthor{Lisa Schlosser, Torsten Hothorn, Achim Zeileis, Reto Stauffer}

@@

  Achim Zeileis \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Achim.Zeileis@R-project.org} \\
  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}

  Reto Stauffer \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"at Innsbruck \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/stauffer/reto-stauffer.html}\\
}

@@

\SweaveOpts{eval = FALSE}

<<echo=FALSE, results=hide>>=
source("oneset.R")
if(file.exists("oneset1.rda")){
  load("oneset1.rda")
} else {
  oneset1 <- sim_oneset(kappa = 7, nobs = 400,
                       seedconst = 7, ntree = 100,
                       formula = y~x1+x2+x3+x4+x5+x6, 
                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
                       forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
                       forest_mtry = 3,
                       type.tree = "ctree",
                       censNO = TRUE,
                       fix.mu = FALSE,
                       fix.sigma = FALSE,
                       mu.sigma.interaction = FALSE,
                       pred_fix = TRUE,
                       pred_fix_x1 = -0.5,
                       pred_fix_x4 = 0,
                       gamboost_cvr = FALSE,
                       eval_disttree = TRUE,
                       eval_distforest = TRUE,
                       eval_gamlss = TRUE,
                       eval_gamboostLSS = FALSE,
                       eval_randomForest = FALSE,
                       eval_cforest = FALSE,
                       mubase = 3,
                       sigmabase = 3)
  save(oneset1, file = "oneset1.rda")
}

if(file.exists("oneset10.rda")){
  load("oneset10.rda")
} else {
  oneset10 <- sim_oneset(kappa = 10, nobs = 400,
                       seedconst = 7, ntree = 100,
                       formula = y~x1+x2+x3+x4+x5+x6, 
                       tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95, 
                       forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
                       forest_mtry = 3,
                       type.tree = "ctree",
                       censNO = TRUE,
                       fix.mu = FALSE,
                       fix.sigma = FALSE,
                       mu.sigma.interaction = FALSE,
                       pred_fix = TRUE,
                       pred_fix_x1 = -0.5,
                       pred_fix_x4 = 0,
                       gamboost_cvr = FALSE,
                       eval_disttree = TRUE,
                       eval_distforest = TRUE,
                       eval_gamlss = TRUE,
                       eval_gamboostLSS = FALSE,
                       eval_randomForest = FALSE,
                       eval_cforest = FALSE,
                       mubase = 3,
                       sigmabase = 3)
  save(oneset10, file = "oneset10.rda")
}
@

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<sim_oneset1_mu, fig=TRUE, echo=FALSE>>=
plot_oneset(oneset1, 
            compare_mu = TRUE, 
            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
            )
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<sim_oneset1_sigma, fig=TRUE, echo=FALSE>>=
plot_oneset(oneset1, 
            compare_sigma_line = TRUE, 
            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
            )
@
\end{subfigure}
\caption{\label{plot_sim_oneset1} ... }
\end{figure}

\begin{figure}[t!]
\begin{subfigure}{0.6\textwidth}
<<sim_oneset10_mu, fig=TRUE, echo=FALSE>>=
plot_oneset(oneset10,
            compare_mu = TRUE,
            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
            )
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.6\textwidth}
<<sim_oneset10_sigma, fig=TRUE, echo=FALSE>>=
plot_oneset(oneset10, 
            compare_sigma_line = TRUE, 
            add_dt = TRUE, add_df = TRUE, add_g = TRUE #, 
            #add_gb = TRUE, add_rf = TRUE, add_cf = TRUE
            )
@
\end{subfigure}
\caption{\label{plot_sim_oneset10} ... }
\end{figure}

@@

<<echo=FALSE, results=hide>>=
if(file.exists("simres.rda")){
  load("simres.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/inst/draft/.R")
  source("gensim.R")
  simres <- gensim(seedconst = 7, nrep = 20, ntree = 100,
                   nsteps = 5, stepsize = 2,
                   formula = y~x1+x2+x3+x4+x5+x6,
                   nobs = 400, testnobs = 200L,
                   tree_minsplit = 25, tree_minbucket = 10, tree_mincrit = 0.95,
                   forest_minsplit = 25, forest_minbucket = 10, forest_mincrit = 0, 
                   forest_mtry = 3,
                   fix.mu = FALSE,
                   fix.sigma = FALSE,
                   mu.sigma.interaction = FALSE,
                   mubase = 1,
                   sigmabase = 3,
                   censNO = TRUE,
                   gamboost_cvr = FALSE,
                   eval_disttree = TRUE,
                   eval_distforest = TRUE,
                   eval_gamlss = TRUE,
                   eval_bamlss = FALSE,
                   eval_gamboostLSS = FALSE,
                   eval_randomForest = FALSE,
                   eval_cforest = FALSE)

  save(simres, file = "simres.rda")
}
@

@@

ncuts <- 100
colR <- colorRampPalette(c("blue","green"))                                       
cols <- colR(ncuts+1)
griddata <- expand.grid(list(x1 = seq(-1, 1, 0.01), x4 = seq(0, 1, 0.2)))
griddata$mu <- levelfun(griddata)
levelplot(mu ~ x1 * x4, data = griddata, xlab = "x1", ylab = "x4", 
          region = TRUE, cuts = ncuts, col.regions = cols)

@@

%Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups. This is the point where the tree structure provides a better solution. However, a single tree can not keep up with the performance of \code{gamlss} when it comes to modeling the smooth functions within the subgroups. Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
%However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
%\\
%Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation. For $\kappa=1$ the performance of all three methods is similar. However, as $\kappa$ increases the RMSE of gamlss increases while the RMSE of trees and forests stays at almost the same level or even decreases sligthly as soon as $\kappa$ is greater or equal 2. The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
%This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
%For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become. Especially regarding the scale parameter the trees are clearly advantageous over the forests.

@@

\SweaveOpts{eval = TRUE}

\newpage
\section{Rain}
In this section precipation data of 30 years is analyzed. The response variable is the meassured total precipiation within 24 hours of each day in July at the station ... .\\
The following methods are applied on this data set:
\begin{itemize}
\item distributional tree
\item distributional forest
\item gamlss
\item gamboostLSS
\item EMOS
\end{itemize}

@@

<<echo=FALSE, results=hide>>=
#### cross validation rain
if(file.exists("rain_Axams.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_Axams.rda")
  load("rain_Axams.rda")
} else {
  #source("~/svn/partykit/pkg/disttree/inst/draft/rain_cross.R")
  source("rain_cross.R")
  library("gamlss.cens")
  gen.cens("NO", type = "left")
  setwd("~/svn/partykit/pkg/disttree/inst/draft/")
  rainres <- rain_cross(stationname = "Axams", 
                        seedconst = 7, ntree = 100,
                        forest_mtry = 27,
                        gamboost_cvr = TRUE,
                        frac = FALSE)
}

rain_rmse <- matrix(0, ncol = ncol(rainres[[1]]$rmse), length(rainres)-1)
rain_ll <- matrix(0, ncol = ncol(rainres[[1]]$ll), length(rainres)-1)
rain_crps <- matrix(0, ncol = ncol(rainres[[1]]$crps), length(rainres)-1)

for(i in 1:(length(rainres)-1)){
  rain_rmse[i,] <- colMeans(rainres[[i]]$rmse, na.rm = TRUE)
  rain_ll[i,] <- colMeans(rainres[[i]]$ll, na.rm = TRUE)
  rain_crps[i,] <- colMeans(rainres[[i]]$crps, na.rm = TRUE)
}

colnames(rain_rmse) <- colnames(rain_ll) <- colnames(rain_crps) <-
  colnames(rainres[[1]]$rmse)

@

\begin{figure}[t!]
\begin{subfigure}{0.3\textwidth}
<<rain_rmse_box, fig=TRUE, echo=FALSE>>=
boxplot(rain_rmse)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.3\textwidth}
<<rain_ll_box, fig=TRUE, echo=FALSE>>=
boxplot(rain_ll)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.3\textwidth}
<<rain_crps_box, fig=TRUE, echo=FALSE>>=
boxplot(rain_crps)
@
\end{subfigure}
\hspace{-1cm}
\begin{subfigure}{0.3\textwidth}
<<rain_crps_box, fig=TRUE, echo=FALSE>>=
boxplot(1 - rain_crps / rain_crps[,7]) 
abline(h = 0, col = 2, lwd = 2)
@
\end{subfigure}
\caption{\label{boxplots_rain}Results 10x10 cross validation on the rain data set (RMSE, log-likelihood and CRPS)\\
left: continuous lines represent the difference between the true and the estimated expected value, dotted lines represent the difference between the observations and the estimated expected value\\
right: continuous lines represent the location parameter, dotted lines the scale parameter}
\end{figure}
